---
title: "Spectral representation and discretisation"
author: "Finn Lindgren"
date: "September 2021, with major updates May/June 2022"
output:
  pdf_document: default
  html_document: default
bibliography: spde10years-combined.bib
header-includes:
  - \usepackage{times,amsmath,amssymb,gensymb,bm}
  - \newcommand{\RR}{\mathbb{R}}
  - \newcommand{\ZZ}{\mathbb{Z}}
  - \newcommand{\s}{\bm{s}}
  - \newcommand{\mmd}{\mathrm{d}}
  - \newcommand{\md}{\,\mmd}
  - \newcommand{\wt}[1]{\widetilde{#1}}
  - \newcommand{\ol}[1]{\overline{#1}}
  - \newcommand{\wh}[1]{\widehat{#1}}
  - \newcommand{\mc}[1]{\mathcal{#1}}
  - \newcommand{\pN}{\textsf{N}}
  - \newcommand{\pCov}{\textsf{Cov}}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=FALSE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(patchwork))
```

The theory discussed in this note follows from the materials found in, among many other places, @Cramer1967, @LindgrenStocProc (available online via the university library), and @LindgrenEtAlStocProc. The materials in the two latter books are later revisions of older teaching materials, and the first is a reprint of the 1967 original.

Key observation:
When the correctly folded spectral density for the spatially discretised process is used, the approximation in the direct spectral representation comes from the frequency *resolution*, not from frequency *truncation*.

Other approximate spectral techniques include the "turning bands" method, that relies
on Monte Carlo integration of the Fourier integrals instead of lattice FFT computations.


## Continuous domains and finite dimensional representations via Fourier theory

The exact spectral representation of the covariance evaluated on a discrete (infinite) lattice is easily derived from the continuous domain representation
$$
\begin{aligned}
R(\s) &= \int_{\RR^d} \exp(i\bm{\omega}\cdot\s) S(\bm{\omega}) \md\bm{\omega} .
\end{aligned}
$$
For simplicity, assume the same lattice spacing $h$ in each direction. The stationary covariance function $R(\s)$ evaluated at lattice points $\bm{k} h$, $k\in\ZZ^d$ is given by
$$
\begin{aligned}
R(\bm{j}h) &= \int_{\RR^d} \exp(i\bm{\omega}\cdot\bm{j} h) S(\bm{\omega}) \md\bm{\omega} \\
 &= \int_{[-\pi/h,\pi/h)^d} \sum_{\bm{k}\in\ZZ^d} \exp(i(\bm{\omega} + 2\pi \bm{k}/h) h \bm{j})
 S(\bm{\omega} + 2\pi \bm{k}/h) \md\bm{\omega}
 \\
 &= \int_{[-\pi/h,\pi/h)^d} \exp(i \bm{\omega} \cdot \bm{j} h)
 \wt{S}(\bm{\omega})
 \md\bm{\omega}, \quad \bm{j}\in \ZZ^d
\end{aligned}
$$
where
$$
\begin{aligned}
\wt{S}(\bm{\omega}) &=\sum_{\bm{k}\in\ZZ^d} S(\bm{\omega} + 2\pi \bm{k}/h),
\quad \bm{\omega}\in [-\pi/h,\pi/h)^d
\end{aligned}
$$
If instead the spatial discretisation should be interpreted as the \emph{cell averages} (which is the more usual case for PDE discretisations and e.g. satellite data, rather than pointwise values), the spectrum is altered by a multiplicative frequency
filter with a squared sinc function:
$$
\begin{aligned}
\wt{S}(\bm{\omega}) &=\sum_{\bm{k}\in\ZZ^d} S(\bm{\omega} + 2\pi \bm{k}/h)
\prod_{l=1}^d
\left\{
\frac{
  \sin[(\omega_l + 2\pi k_l/h) h / 2]
}{
(\omega_l + 2\pi k_l/h) h / 2
}
\right\}^2,
\quad \bm{\omega}\in [-\pi/h,\pi/h)^d .
\end{aligned}
$$


### Sampling, in theory

According to the spectral representation theory for processes with continuous
spectra, we have
$$
\begin{aligned}
x(\s) &= \int_{\RR^d} \exp(i\bm{\omega}\cdot\s) S(\bm{\omega})^{1/2} \md Z(\bm{\omega}),
\quad \s\in\RR^d ,
\end{aligned}
$$
where where $\mmd Z(-\bm{\omega})=\ol{\mmd Z(\bm{\omega})}$, $\textsf{Cov}(\mmd Z(\bm{\omega}),\mmd Z(\bm{\omega}'))= \delta(\bm{\omega}-\bm{\omega}')\md\bm{\omega}$ for $\bm{\omega}\in\RR^d$.

With the above theory, for sampling, the integral again reduces to a finite domain,
but one needs to construct a spectral white noise process
$\mmd Z(\bm{\omega})$ with complex conjugate symmetry:
\begin{align*}
x(\bm{j}h)
 &= \int_{[-\pi/h,\pi/h)^d} \exp(i \bm{\omega} \cdot \bm{j} h)
\wt{S}(\bm{\omega})^{1/2} \md Z(\bm{\omega}), \quad \bm{j}\in \ZZ^d,
\end{align*}
where $\mmd Z(-\bm{\omega})=\ol{\mmd Z(\bm{\omega})}$, $\textsf{Cov}(\mmd Z(\bm{\omega}),\mmd Z(\bm{\omega}'))= \delta(\bm{\omega}-\bm{\omega}')\md\bm{\omega}$ for $\bm{\omega}\in [-\pi/h,\pi/h)^d$.
This can in principle be discretised with a
lattice of frequencies in much the same way as a basic integral approximation
method that can be used to compute the deterministic integral for the covariance function,
with noise variances equal to the cell area/volume of each frequency lattice point.
However, the complex conjugacy can be tricky to work with in practice. In the next section
we will solve this issue, and provide a fast computational method.

## Practical numerical evaluation with FFT

Above we let the spacings $h$ be the same in all dimensions. For greater generality,
we now define discrete spacings $\bm{h}=\{h_1,\dots,h_d\}$.
Let the number of evaluation points in each dimension be given by
$\bm{n}=\{n_1,\dots,n_d\}$. For the cell interpretation of the spatial
discretisation, the domain with for the different directions are given by 
$\bm{L}=\bm{n}\odot\bm{h}$.
Define the continuous spectral domain
$\mc{I}=\prod_{l=1}^d [-\pi/h_l,\pi/h_l)$, a semi-symmetric index space as
$\mc{K}_0=\prod_{l=1}^d \{-n_l/2,\dots,-1,0,1,2,\dots,n_l/2-1\}$, and
and a non-negative index space
$\mc{K}_+=\prod_{l=1}^d \{0,1,2,\dots,n_l-1\}$, so that if $\bm{k}\in\mc{K}_0$,
then $\bm{k}\bmod \bm{n}\in\mc{K}_+$.

We want to evaluate
$$
\begin{aligned}
x(\bm{j}\odot\bm{h}) &= \int_{\mc{I}} \exp(i\bm{\omega}\cdot\bm{j} \odot\bm{h}) \wh{x}(\bm{\omega}) \md\bm{\omega}
\end{aligned}
$$
for all $\bm{j}\in\bm{K}_+$ or all $\bm{j}\in\bm{K}_0$.

The (inverse) FFT in `fftwtools` is a by default unnormalised sum,
$$
\begin{aligned}
x_{\bm{j}} &= \sum_{\bm{k}\in\mc{K}_+} \exp(i 2\pi\bm{k}\cdot\bm{j} \oslash \bm{n}) \wh{x}_{\bm{k}}
,\quad\bm{j}\in\mc{K}_+ .
\end{aligned}
$$
so we need to make a Riemann sum approximation of the integral and write it on this form.

Define a discretisation of the $\bm{\omega}$ space via
$$
(\bm{\omega}_{\bm{k}})_l = \frac{\pi k_l}{h_l n_l/2} = \frac{2\pi k_l}{h_l n_l}, \quad k_l = -n_l/2,\dots,n_l/2-1
$$
Then
$$
\begin{aligned}
x(\bm{j}\odot\bm{h}) &\approx \sum_{\{k_l=-n_l/2\}}^{\{n_l/2-1\}} \exp(i2\pi\bm{k}\cdot\bm{j} \oslash \bm{n})
\wh{x}(2\pi\bm{k}\oslash(\bm{n}\odot\bm{h})) \prod_{l=1}^d \frac{2\pi}{h_l n_l}
\end{aligned}
$$
For the negative $\bm{k}$ indices, we can add a shift by $\bm{n}$ without altering the result (if we also adjust $\wh{x}$ accordingly), since $\exp(i2\pi\bm{n}\cdot\bm{j}\oslash\bm{n})=\exp(i2\pi\bm{j})=1$ for all $\bm{j}$.
We can therefore compute an approximation of the integral by letting
$$
\wh{x}_{\bm{k} \bmod \bm{n}}=
\wh{x}(2\pi\bm{k}\oslash(\bm{n}\odot\bm{h}))\prod_{l=1}^d\frac{2\pi}{h_l n_l},
\quad\bm{k}\in\mc{K}_0.
$$


### Covariance evaluation

To compute an approximation of the covariance function, we define
$$
\sigma_{\bm{k}}^2 = \wt{S}(2\pi\bm{k}\oslash(\bm{n}\odot\bm{h})) \prod_{l=1}^d\frac{2\pi}{h_l n_l},
\quad \bm{k}\in\mc{K}_0,
$$
and let
$$
\wh{x}_{\bm{k} \bmod \bm{n}} = \sigma_{\bm{k}}^2,
\quad \bm{k}\in\mc{K}_0 .
$$
Then $R(\bm{j}\odot\bm{h})\approx x_{\bm{j}\bmod\bm{n}}$ for $\bm{j}\in\mc{K}_0$.

### Process/field simulation

For simulation, the relation $\mmd Z(-\bm{\omega})=\ol{\mmd Z(\bm{\omega})}$ makes
direct construction of the discretised integrand difficult, as we need to keep
track of the complex conjugate symmetry. A practical alternative can be found in
Appendix B.4 of [@LindgrenStocProc], where it is shown that if the complex conjugate
symmetry is ignored, the result of the (inverse) Fourier transform gives
a sample from the process as its real part, and a corresponding sample from the
process _envelope_ as the norms of the absolute values.

We here present a similar method that is easier to show gives the correct result.
Let $(\bm{\omega}_{\bm{k}})_l=\frac{2 \pi k_l}{h_l n_l}$ for $k_l=-n_l/2,1,\dots,n_l/2-1$.
Define
$$
\sigma_{\bm{k}}^2 = \wt{S}(\bm{\omega}_{\bm{k}}) \prod_{l=1}^d\frac{2\pi}{h_l n_l},
\quad \bm{k}\in\mc{K}_0,
$$

Let $\wh{z}_{\bm{k}}$ be iid complex valued $\pN(0,1)$ random variables, and let
$$
\wh{x}_{\bm{k}} =
\wh{z}_{\bm{k}} \sigma_{\bm{k}},
\quad \bm{k}\in\mc{K}_0 .
$$
Then the real part of the Fourier transformed values, $x(\bm{j}\odot\bm{h})=\Re[x_{\bm{j}}]$ form an approximate sample from the process model, $\bm{j}\in\mc{K}_+$.

[and $|x_{\bm{j}}|$ is the corresponding sample of the process envelope, for
$\bm{j}\in\mc{K}_+$.]

#### Proof

Let $a_{\bm{k}}$ and $a_{\bm{k}}$ be the real and imaginary parts of $\wh{z}_{\bm{k}}$.
Define
$c_{\bm{k}\cdot\bm{j}}=\cos(\bm{\omega}_{\bm{k}}\cdot\bm{j} \odot \bm{h})$ and
$s_{\bm{k}\cdot\bm{j}}=\sin(\bm{\omega}_{\bm{k}}\cdot\bm{j} \odot \bm{h})$.
Then
$$
\begin{aligned}
x_{\bm{j}} &=
\sum_{\bm{k}\in\mc{K}_0}
\exp(i\bm{\omega}_{\bm{k}}\cdot\bm{j} \odot \bm{h}) (a_{\bm{k}} + i b_{\bm{k}})\sigma_{\bm{k}}
\\&=
\sum_{\bm{k}\in\mc{K}_0}
(c_{\bm{k}\cdot\bm{j}} + i s_{\bm{k}\cdot\bm{j}}) (a_{\bm{k}} + i b_{\bm{k}})\sigma_{\bm{k}}
\\&=
\sum_{\bm{k}\in\mc{K}_0}
(c_{\bm{k}\cdot\bm{j}} a_{\bm{k}} - s_{\bm{k}\cdot\bm{j}} b_{\bm{k}})\sigma_{\bm{k}}
+
i \sum_{\bm{k}\in\mc{K}_0}
(c_{\bm{k}\cdot\bm{j}} b_{\bm{k}} + s_{\bm{k}\cdot\bm{j}} a_{\bm{k}})\sigma_{\bm{k}} .
\end{aligned}
$$
The covariance for the real part is given by
$$
\begin{aligned}
\pCov[\Re(x_{\bm{j}}),\Re(x_{\bm{j}'})] &=
\sum_{\bm{k}\in\mc{K}_0}
\pCov[
(c_{\bm{k}\cdot\bm{j}} a_{\bm{k}} - s_{\bm{k}\cdot\bm{j}} b_{\bm{k}})\sigma_{\bm{k}},
(c_{\bm{k}\cdot\bm{j}'} a_{\bm{k}} - s_{\bm{k}\cdot\bm{j}'} b_{\bm{k}})\sigma_{\bm{k}}]
\\&=
\sum_{\bm{k}\in\mc{K}_0}
(c_{\bm{k}\cdot\bm{j}}c_{\bm{k}\cdot\bm{j}'} + s_{\bm{k}\cdot\bm{j}}s_{\bm{k}\cdot\bm{j}'})\sigma_{\bm{k}}^2
\\&=
\sum_{\bm{k}\in\mc{K}_0}
c_{\bm{k}\cdot(\bm{j}'-\bm{j})} \sigma_{\bm{k}}^2 .
\end{aligned}
$$
where the last step follows from the trigonometric identity for the cosine of the difference between two angles.
By considering the definition of $\sigma_{\bm{k}}^2$, we see that the resulting sum is a Riemann sum approximation of the cosine integral
$$
\begin{aligned}
\int_{\prod_{l=0}^d [-\pi/h_l,\pi/h_l)} &\cos(\bm{\omega}\cdot (\bm{j}'-\bm{j}) \odot\bm{h}) \wt{S}(\bm{\omega}) \mathrm{d}\bm{\omega}
=
\int_{\prod_{l=0}^d [-\pi/h_l,\pi/h_l)} \cos(\bm{\omega}\cdot(\bm{j}'-\bm{j}) \odot\bm{h}) \wt{S}(\bm{\omega})
\mathrm{d}\bm{\omega}
\\&=
\int_{\prod_{l=0}^d [-\pi/h_l,\pi/h_l)} \exp(i\bm{\omega}\cdot(\bm{j}'-\bm{j}) \odot\bm{h}) \wt{S}(\bm{\omega}) \mathrm{d}\bm{\omega}
\end{aligned}
$$
where in the final step, the integral of the $\sin(i\bm{\omega}\cdot(\bm{j}'-\bm{j}) \odot\bm{h})$ term from $\exp()$ vanishes due to symmetry.
This shows that we have the spectral representation of the covariance function, $R[\bm{h}\odot(\bm{j}'-\bm{j})]$, for the given grid resolution (with $\wt{S}$ incorporating the required spectral folding, etc).

Note: The imaginary part of $x_{\bm{j}}$ leads to the same covariance calculations as the real part. However,
the cross-covariance between the real imaginary parts is non-zero, so the real and imaginary parts of the sample are not independent. This is because even though the $s_{\bm{k}\cdot(\bm{j}'-\bm{j})}$ terms cancel out pairwise for $\bm{k}$ and $-\bm{k}$, the indices where one or more $k_l=-n_l/2$ have no corresponding $k_l=n_l/2$ in $\mc{K}_0$.
Thus, in order to generate two independent samples in a single complex FFT evaluation, one needs to set such "lower boundary" $\sigma_{\bm{k}}$ values to zero. This slightly reduces the accuracy of the sampling, but since it lowers the computational cost by a factor 2, and only affects frequencies that typically have small contribution, this may be a preferable approach in practice.

### A note on resolution

When choosing the frequency resolution for the Fourier integral discretisation,
we are constrained by the desired space resolution $\bm{h}$, but we are
 free to choose the number of intervals in each dimension, $\bm{n}$. However,
 as a by-product of the FFT construction, this also dictates that the number of evaluation
 points in space will _also_ be $\bm{n}$. Thus, even if we only desire to evaluate
 the covariance or simulation at a small number of spatial points, we still need
 to use enough discretisation points to obtain an accurate Fourier integral approximation.
 In practice, this means that we might only use a small portion of the computed results.
 

## Transformation code

```{r S2C,code=readLines("S2C.R")}
source("S2C.R")
```


## 2D example

```{r}
S_fun <- function(omega) {
  rho <- 10
  exp(-rowSums(omega^2) / rho^2 / 2) / rho^2 / (2 * pi)
}
n <- c(256, 256)
L <- c(1, 1)
h <- L / n
x_ <- make_x(n, L)
omega_ <- make_omega(n, L)
omega <- as.matrix(expand.grid(omega_))
S_truncated <- S_fun(omega = omega)
S_folded <- fold_spectrum(omega = omega, S_fun = S_fun, h)
S_folded_cell <- fold_spectrum(omega = omega, S_fun = S_fun, h, pointwise = FALSE)
```

```{r}
S_df <-
    cbind(
      as.data.frame(as.matrix(expand.grid(omega_))),
      data.frame(
        Truncated = S_truncated,
        Folded = S_folded,
        "Folded_cell" = S_folded_cell
      )
    ) %>%
      pivot_longer(
        cols = c("Truncated", "Folded", "Folded_cell"),
        names_to = "Type", values_to = "Value"
      )

C_df <-
    cbind(
      as.data.frame(as.matrix(expand.grid(x_))),
      data.frame(
        Truncated = as.vector(S2C(S_truncated, n, h)),
        Folded = as.vector(S2C(S_folded, n, h)),
        "Folded_cell" = as.vector(S2C(S_folded_cell, n, h))
      )
    ) %>%
      pivot_longer(
        cols = c("Truncated", "Folded", "Folded_cell"),
        names_to = "Type", values_to = "Value"
      )
```

```{r}
ggplot(S_df) +
  geom_raster(aes(w1, w2, fill = Value)) +
  facet_grid(vars(Type)) +
  xlab("omega1") + ylab("omega2") +
  labs(fill = "Spectral density")
```

```{r}
ggplot(C_df) +
  geom_raster(aes(x1, x2, fill = Value)) +
  facet_grid(vars(Type)) +
  xlab("x1") + ylab("x2") +
  labs(fill = "Covariance")
```

```{r}
S_df %>% group_by(Type) %>% summarise(N = sum(Value > 0))
```

## 1D example

```{r}
#S_fun <- function(omega) {
#  rho <- 2
#  exp(-rowSums(omega^2) * rho^2 / 2) / sqrt(2 * pi) * rho
#}
S_fun <- function(omega) {
  nu <- 1.5
  rho <- 3 / sqrt(8 * nu)
  v <- rho^(2 * nu) * gamma(nu) / gamma(nu + 0.5) / sqrt(4 * pi)
  rho^(2 * nu + 1) / (2 * pi) /
    (1 + rowSums(omega^2) * rho^2)^(nu + 0.5)
}
#S_fun <- function(omega) {
#  rho <- 0.1
#  1/(1-0.9*2*(rowSums(omega^2) * rho^2)^2+(rowSums(omega^2) * rho^2)^4)^2
#}
n <- c(512)
L <- c(20)
h <- L / n
x_ <- make_x(n, L)
omega_ <- make_omega(n, L)
omega <- as.matrix(expand.grid(omega_))
S_truncated <- fold_spectrum(omega = omega, S_fun = S_fun, h, fold = FALSE)
S_folded <- fold_spectrum(omega = omega, S_fun = S_fun, h)
S_folded_cell <- fold_spectrum(omega = omega, S_fun = S_fun, h, pointwise = FALSE)


S_df <-
    cbind(
      as.data.frame(as.matrix(expand.grid(omega_))),
      data.frame(
        Truncated = S_truncated,
        Folded = S_folded,
        "Folded_cell" = S_folded_cell
      )
    ) %>%
      pivot_longer(
        cols = c("Truncated", "Folded", "Folded_cell"),
        names_to = "Type", values_to = "Value"
      )

C_df <-
    cbind(
      as.data.frame(as.matrix(expand.grid(x_))),
      data.frame(
        Truncated = as.vector(S2C(S_truncated, n, h)),
        Folded = as.vector(S2C(S_folded, n, h)),
        "Folded_cell" = as.vector(S2C(S_folded_cell, n, h))
      )
    ) %>%
      pivot_longer(
        cols = c("Truncated", "Folded", "Folded_cell"),
        names_to = "Type", values_to = "Value"
      )
```

```{r}
ggplot(S_df, aes(w1, Value)) +
  geom_line(aes(col = Type)) +
  xlab("omega") + ylab("Spectral density")
```

```{r}
ggplot(C_df, aes(x1, Value)) +
  geom_line(aes(col = Type)) +
  coord_cartesian(xlim = c(0, L / 2)) +
  xlab("x") + ylab("Covariance")
```

```{r}
S_df %>% group_by(Type) %>% summarise(N = sum(Value > 0))
```

## Sampling example

```{r}
n_sampling <- n * 2
x_ <- make_x_sampling(n_sampling, h)
omega_ <- make_omega_sampling(n_sampling, h)
omega <- as.matrix(expand.grid(omega_))
S_truncated <- S_fun(omega = omega)
S_folded <- fold_spectrum(omega = omega, S_fun = S_fun, h)
S_folded_cell <- fold_spectrum(omega = omega, S_fun = S_fun, h, pointwise = FALSE)
samples_df <-
    cbind(
      as.data.frame(as.matrix(expand.grid(x_))),
      data.frame(
        Truncated = as.vector(S2sample(S_truncated, n_sampling, h, seed = 123L)),
        Folded = as.vector(S2sample(S_folded, n_sampling, h, seed = 123L)),
        "Folded_cell" = as.vector(S2sample(S_folded_cell, n_sampling, h, seed = 123L))
      )
    ) %>%
      pivot_longer(
        cols = c("Truncated", "Folded", "Folded_cell"),
        names_to = "Type", values_to = "Value"
      )
```

```{r}
ggplot(samples_df %>% filter(x1 < L), aes(x1)) +
  geom_line(aes(y = Re(Value), col = Type, lty="Real part")) +
  geom_line(aes(y = Im(Value), col = Type, lty="Imag part")) +
  coord_cartesian(xlim = c(0, L)) +
  xlab("x") + ylab("Sample")
```
(for long correlation range, the three versions are superimposed, since they used the same random seed, and the spectra are almost identical even after folding/truncation)

## References


